miniKanren with fair search strategies

### 2019-05-02 Paper/paper.tex
state->answer

The syntax of a programming language should reflect its semantics. When writing 
a \conde expression in miniKanren, a programmer would expect all clauses share 
the same chance of being explored, as these clauses are written in 
parallel. The existing search strategy, interleaving DFS (DFS$_\textrm{i}$), however, 
prioritize its clauses by the order how they are written down. Similarly, when 
a \conde is followed by another goal conjunctively, a programmer would expect 
states in parallel share the same chance of being explored. Again, the answers 
by DFS$_\textrm{i}$ is different from the expectation. We have devised three new search 
strategies that have different level of fairness in \disj and \conj.

@@

We demonstrate the aspects (\disj{} or \conj{}) and levels (unfair, 
almost-fair, or fair) of fairness by running queries about \texttt{repeato}, a 
relational definition that relates a term \texttt{x} with a non-empty list 
whose elements are \texttt{x} (Fig.~\ref{repeato}).

@@

In the following program, the three \conde clauses differ in a trivial way. So 
we expect lists of each letter constitute $1/4$ of the answer list. However, 
DFS$_\textrm{i}$, the current search strategy, gives us much more lists of \texttt{a}s than 
lists of other letters. And some letters (e.g. lists of \texttt{c}) are rarely 
found. The situation would be worse if we add more \conde clauses.

@@

We borrow the definition of \emph{fair} \disj{} from \cite{seres1999algebra}: 
search strategies with \emph{fair} \disj{} should allocate resource evenly 
among disjunctive goals. Running the same program with DFS$_\textrm{f}$ and BFS give the 
following result.

@@

We conceive a middle place between fair and unfair -- search strategies with 
\emph{almost-fair} \disj{} should allocate resource so evenly among disjunctive 
goals\citep{seres1999algebra} that the the maximal ratio of resource is bounded 
by a constant. Our new search strategy, DFS$_\textrm{bi}$, has almost-fair \disj{}. It is 
fair when the number of goals is a power of 2, otherwise, some goals are 
allocated twice more resource than the others. In the previous example, it 
gives the same result. And in the following example, where the \conde has 5 
clause, the clauses of \texttt{b}, \texttt{c}, and \texttt{d} are allocated 
more resource.

@@

A more interesting situation is when the first conjunctive goal produces 
infinite many states. Consider the following example, a naive specification of 
fair \conj{} might require search strategies to produce all sorts of singleton 
lists, but no longer ones, which makes the strategies \emph{incomplete}.

@@

Our solution is requiring a search strategy with \emph{fair} \conj{} to 
package states in bags, where each bag contains finite states, and to allocate 
resource evenly among search spaces derived from states in the same bag. The 
way to package depends on search strategy. And how to allocate resource among 
search space related to different bags is unspecified. Our definition of fair 
\conj{} is orthogonal with completeness. For example, a naively fair strategy 
is fair but not complete, while BFS is fair and complete.

BFS packages states by their costs. The \emph{cost} of a state is its 
depth in the search tree (i.e. the number of calls to relational definitions 
required to find them) \citep{seres1999algebra}. In the following example, the 
cost of an answer is equal to the length of the inner lists plus the length of 
the outer list. Noted that BFS produces answers in increasing order of cost.

@@

The implementation can be improved in two aspects. First, as we mentioned in 
section 2.2, our BFS bag states by their cost. However, in this implementation, 
it is unclear where this information is recorded. Second, 
\texttt{append-inf/fair} is extravagant in memory usage. It makes 
$O(n+m)$ new \texttt{cons} cells every time, where $n$ and $m$ are the 
``length''s of input search spaces. We address these issues in the first 
subsection.

@@

As we mentioned in section 2.2, our BFS bag states by their cost. The bagging 
information is recorded subtly -- the \texttt{car}s of a search space have cost 
0 (i.e. they are in the same bag), and the costs of states in thunk are 
computed recursively then increased by one. It is even more subtle that
\texttt{append-inf/fair} and the \texttt{append-map-inf/fair} respects the cost 
information. We make these facts more obvious by changing the type of search 
space, modifying related function definitions, and introducing a few more 
functions.

The new type is a pair whose \texttt{car} is a list of state (the bag), and 
whose \texttt{cdr} is either a \texttt{\#{}f} or a thunk returning a search 
space. A falsy \texttt{cdr} means the search space is obviously finite.

Functions related to the pure subset are listed in Fig.~\ref{BFS-opt} (the 
others in Fig.~\ref{BFS-opt-cont}). We take them apart because they are 
compared with Seres's implementation later. The first three functions in 
Fig.~\ref{BFS-opt} are search space constructor. \texttt{none} makes an empty 
search space. \texttt{unit} makes an space from one state. \texttt{step} makes 
a space from a thunk. The remaining functions do the same thing as before.

@@

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \texttt{elim}, takes a space \texttt{s-inf} and two 
continuations \texttt{ks} and \texttt{kf}. When \texttt{s-inf} contains some 
states, \texttt{ks} is called with the first state and the rest space. 
Otherwise, \texttt{kf} is called with no argument. Here `s' and `f' means 
`succeed' and `fail' respectively. This function is an eliminator of search 
space, whence the name. The remaining functions do the same thing as before.

@@

In this section, we compare the pure subset of our optimized BFS with the BFS 
found in \citep{seres1999algebra}. We focus on the pure subset because 
Silvija's system is pure. Her system represents search spaces with streams of 
lists of states, where each list is a bag.

@@

In general, only DFS$_\textrm{i}$ and DFS$_\textrm{bi}$ constantly perform well. DFS$_\textrm{f}$ is just as 
efficient in all benchmarks but \texttt{very-recursiveo}. Both BFS have obvious 
overhead in many cases. Among the three variants of DFS (they all have unfair 
\conj{}), DFS$_\textrm{f}$ is most resistant to clause permutation, followd by DFS$_\textrm{bi}$ then 
DFS$_\textrm{i}$. Among the two implementation of BFS, ours constantly performs as well or 
better. Interestingly, every strategies with fair \disj{} suffers in 
\texttt{very-recursiveo} and DFS$_\textrm{f}$ performs well elsewhere. Therefore, this 
benchmark might be a special case. Fair \conj{} imposes overhead constantly 
except in \texttt{appendo}. The reason might be that strategies with fair 
\conj{} tend to keep more intermediate states in the memory.

>>>>

The syntax of a programming language should reflect its semantics. When writing 
a \conde expression in miniKanren, a programmer would expect all clauses share 
the same chance of being explored, as these clauses are written in 
parallel. The existing search strategy, interleaving DFS (DFS$_\textrm{i}$), however, 
prioritize its clauses by the order how they are written down. Similarly, when 
a \conde is followed by another goal conjunctively, a programmer would expect 
answers in parallel share the same chance of being explored. Again, the answers 
by DFS$_\textrm{i}$ is different from the expectation. We have devised three new search 
strategies that have different level of fairness in \disj and \conj.

@@

In this section, we demonstrate the aspects (\disj{} or \conj{}) and levels
(unfair, almost-fair, or fair) of fairness by running queries about
\texttt{repeato}, a 
relational definition that relates a term \texttt{x} with a non-empty list 
whose elements are \texttt{x} (Fig.~\ref{repeato}).

@@

Given the following program, it is natural to expect lists of each letter
constitute $1/4$ in the answer. However, DFS$_\textrm{i}$, the current search
strategy, results in much more lists of \texttt{a}s than lists of other
letters. And some letters, e.g. \texttt{c} and \texttt{d}, are rarely seen.
The situation would be exacerbated if \conde contains more clauses.

@@

We borrow the definition of \emph{fair} \disj{} from \cite{seres1999algebra}: 
search strategies with \emph{fair} \disj{} should allocate resources evenly 
among disjunctive goals. Running the same program with DFS$_\textrm{f}$ and BFS give the 
following result.

@@

Now we are in a middle place between fair and unfair -- search strategies
with  \emph{almost-fair} \disj{} should allocate resources so evenly among
disjunctive  goals\citep{seres1999algebra} that the the maximal ratio of
resources is bounded by a constant. Our new search strategy,
DFS$_\textrm{bi}$, has almost-fair \disj{}. It is
fair when the number of goals is a power of 2, otherwise, some goals are 
allocated twice more resources than the others. In the previous example,
DFS$_\textrm{bi}$
gives the same result. And in the following example, where the \conde has 5 
clause, the clauses of \texttt{b}, \texttt{c}, and \texttt{d} are allocated 
more resources.

@@

A more interesting situation is when the first conjunctive goal produces 
infinite many answers. Consider the following example, a naive specification of 
fair \conj{} might require search strategies to produce all sorts of singleton 
lists, but no longer ones, which makes the strategies \emph{incomplete}.

@@

Our solution is requiring a search strategy with \emph{fair} \conj{} to 
package answers in bags, where each bag contains finite answers, and to allocate 
resources evenly among search spaces derived from answers in the same bag. The 
way to package depends on search strategy. And how to allocate resources among 
search space related to different bags is unspecified. Our definition of fair 
\conj{} is orthogonal with completeness. For example, a naively fair strategy 
is fair but not complete, while BFS is fair and complete.

BFS packages answers by their costs. The \emph{cost} of a answer is its 
depth in the search tree (i.e. the number of calls to relational definitions 
required to find them) \citep{seres1999algebra}. In the following example, the 
cost of an answer is equal to the length of the inner lists plus the length of 
the outer list. Noted that BFS produces answers in increasing order of cost.

@@

The implementation can be improved in two aspects. First, as we mentioned in 
section 2.2, our BFS bag answers by their cost. However, in this implementation, 
it is unclear where this information is recorded. Second, 
\texttt{append-inf/fair} is extravagant in memory usage. It makes 
$O(n+m)$ new \texttt{cons} cells every time, where $n$ and $m$ are the 
``length''s of input search spaces. We address these issues in the first 
subsection.

@@

As we mentioned in section 2.2, our BFS bag answers by their cost. The bagging 
information is recorded subtly -- the \texttt{car}s of a search space have cost 
0 (i.e. they are in the same bag), and the costs of answers in thunk are 
computed recursively then increased by one. It is even more subtle that
\texttt{append-inf/fair} and the \texttt{append-map-inf/fair} respects the cost 
information. We make these facts more obvious by changing the type of search 
space, modifying related function definitions, and introducing a few more 
functions.

The new type is a pair whose \texttt{car} is a list of answers (the bag), and 
whose \texttt{cdr} is either a \texttt{\#{}f} or a thunk returning a search 
space. A falsy \texttt{cdr} means the search space is obviously finite.

Functions related to the pure subset are listed in Fig.~\ref{BFS-opt} (the 
others in Fig.~\ref{BFS-opt-cont}). We take them apart because they are 
compared with Seres's implementation later. The first three functions in 
Fig.~\ref{BFS-opt} are search space constructor. \texttt{none} makes an empty 
search space. \texttt{unit} makes an space from one answer. \texttt{step} makes 
a space from a thunk. The remaining functions do the same thing as before.

@@

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \texttt{elim}, takes a space \texttt{s-inf} and two 
continuations \texttt{ks} and \texttt{kf}. When \texttt{s-inf} contains some 
answers, \texttt{ks} is called with the first answer and the rest space. 
Otherwise, \texttt{kf} is called with no argument. Here `s' and `f' means 
`succeed' and `fail' respectively. This function is an eliminator of search 
space, whence the name. The remaining functions do the same thing as before.

@@

In this section, we compare the pure subset of our optimized BFS with the BFS 
found in \citep{seres1999algebra}. We focus on the pure subset because 
Silvija's system is pure. Her system represents search spaces with streams of 
lists of answers, where each list is a bag.

@@

In general, only DFS$_\textrm{i}$ and DFS$_\textrm{bi}$ constantly perform well. DFS$_\textrm{f}$ is just as 
efficient in all benchmarks but \texttt{very-recursiveo}. Both BFS have obvious 
overhead in many cases. Among the three variants of DFS (they all have unfair 
\conj{}), DFS$_\textrm{f}$ is most resistant to clause permutation, followd by DFS$_\textrm{bi}$ then 
DFS$_\textrm{i}$. Among the two implementation of BFS, ours constantly performs as well or 
better. Interestingly, every strategies with fair \disj{} suffers in 
\texttt{very-recursiveo} and DFS$_\textrm{f}$ performs well elsewhere. Therefore, this 
benchmark might be a special case. Fair \conj{} imposes overhead constantly 
except in \texttt{appendo}. The reason might be that strategies with fair 
\conj{} tend to keep more intermediate answers in the memory.
